---
title: "Practical Machine Learning"
author: "Kelvin Han"
date: "17 August 2015"
output: html_document
---

### Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

### Data Source 

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

### Data Exploration
From the summary, it is observed that 
a) X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp and new_window are redundant
b) there are blank and #DIV/0! values
```{r}
DataExploration <- read.csv("pml-training.csv")
summary(DataExploration)
```

### Data Cleaning
Read the two csv files into two data frames and defining “NA” and “#DIV/0!” as na.strings. The following columns are removed.
a) 6 redundant columns: X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp and new_window  
b) 4 columns with number of NA values at 100%  
c) 96 columns with number of NA values > 90%
```{r}
trainingData <- read.csv("pml-training.csv",header = TRUE, na.strings = c("NA", "#DIV/0!"))
testData <- read.csv("pml-testing.csv",header = TRUE, na.strings = c("NA", "#DIV/0!"))

trainingData <- Filter(function(x)!any(is.na(x)), trainingData)
trainingData$X <- NULL
trainingData$user_name <- NULL
trainingData$raw_timestamp_part_1 <- NULL
trainingData$raw_timestamp_part_2 <- NULL
trainingData$cvtd_timestamp <- NULL
trainingData$new_window <- NULL
trainingData$num_window <- NULL

testData <- Filter(function(x)!any(is.na(x)), testData)
testData$X <- NULL
testData$user_name <- NULL
testData$raw_timestamp_part_1 <- NULL
testData$raw_timestamp_part_2 <- NULL
testData$cvtd_timestamp <- NULL
testData$new_window <- NULL
testData$num_window <- NULL

dim(trainingData)
dim(testData)
```

### Libraries Required
Using Caret, RandomForest and doMC (for MAC)
```{r}
library(caret)
library(randomForest)
library(doMC)
registerDoMC(cores = 2)
```


### Datasets Preparation
Split the training data into a training (60%) and validation (40%) dataset.
```{r}
set.seed(12345)
inTrain <- createDataPartition(y=trainingData$classe,p=0.6, list=FALSE)
trainingSet <- trainingData[inTrain,]
testingSet <- trainingData[-inTrain,]
```

### Prediction Model Training
Using various models, we shortlist the five highest accuracy models
RandomForest 0.99		
Single C5.0 Ruleset			0.96
Single C5.0 Tree			0.95
Support Vector Machines with Radial Basis Function Kernel 0.92
Quadratic Discriminate Analysis 	0.89
```{r}
system.time(modelFitRF <- train(classe ~., data=trainingSet, method="rf", trControl=trainControl (method="cv",number=5), prox=TRUE, allowParallel=TRUE))
system.time(modelFitC5R <- train(classe ~., data=trainingSet, method="C5.0Rules"))
system.time(modelFitC5T <- train(classe ~., data=trainingSet, method="C5.0Tree"))
system.time(modelFitSVM <- train(classe ~., data=trainingSet, method="svmRadial", trControl=trainControl (method="cv",number=5), prox=TRUE, allowParallel=TRUE))
system.time(modelFitQDA <- train(classe ~., data=trainingSet, method="qda"))

predictModelRF <- predict(modelFitRF,testingSet)
predictModelSVM <- predict(modelFitSVM,testingSet)
predictModelC5R <- predict(modelFitC5R,testingSet)
predictModelC5T <- predict(modelFitC5T,testingSet)
predictModelQDA <- predict(modelFitQDA,testingSet)
as.numeric(confusionMatrix(predictModelRF,testingSet$classe)$overall[1])
as.numeric(confusionMatrix(predictModelC5R,testingSet$classe)$overall[1])
as.numeric(confusionMatrix(predictModelC5T,testingSet$classe)$overall[1])
as.numeric(confusionMatrix(predictModelSVM,testingSet$classe)$overall[1])
as.numeric(confusionMatrix(predictModelQDA,testingSet$classe)$overall[1])
```

Confusion Matrix for the best model (Random Forest) based on accuracy
```{r}
confusionMatrix(predictModelRF,testingSet$classe)
```

The expected out of sample error is expected to be very small (<1%)
```{r}
prediction <- predict(modelFitRF,testData)
```

### Files generation for submission
Generate files for submission
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(prediction)
```